---
output:
  html_document: default
  pdf_document:
    number_sections: yes
---

# SCR live coding Sept 7, 2019


## Session 1: Objects and Functions


### objects a.k.a. variables

```{r, error = TRUE}
a <- 9
b <- TRUE
c <- "two"
d <- UNKNOWN # does not work: R has some built-in 'objects', such as numbers, TRUE and FALSE
# notice the colouring in an IDE environment, also e.g. tabbing
```





### using functions (and operators)


```{r}
sqrt(a)
print(b) 
```

`print(b)` is the same as simply entering 'b' in your console


Let's take a look at

```{r}
b * 2
```

Apparantly R has some built-in (default) behaviours, TRUE somehow becomes a 1. How about FALSE?

```{r}
d <- FALSE 
d * 2
```

Usually these default behaviours are convenient, sometimes confusing! Let's get Back to the lecture...



\newpage 
## Session 2: Objects and Functions


### creating your own function

```{r}
square <- function(x){
  squared <- x * x
  return(squared)
}
```

Did you see we use the assignment arrow? Actually, functions are also variables / objects. Because we (usually) use them functions in such a fundamentally different way, especially for now, we do not view them as variables or objects.


```{r, error = TRUE}
square(a)
sqrt(square(a))
square(b) # error!
```



### side effects and errors.. NA / NaN / Null?

About `Na`

```{r, error = TRUE}
as.numeric("four")
as.numeric(c("3", "2", "...", "1"))
```

About `NaN`
```
1/0 - 1/0
c(1, 2, NA, 4)
```

About `NULL`

```{r, eval = FALSE}
a <- plot("")
c(3, 2, NULL, 1)
```

Some checks: 
```{r}
is.na(NA)
is.null(NULL)
is.null(NA)
```

and tricky ones:

```{r}
is.na(NULL) # the answer is a logical vector (TRUE / FALSE) of length 0. E.g. an empty vector
is.null(logical(0)) # an empty vector != nothingness
```


Meanwhile on your hard drive:
```{r}
object.size(NULL)
object.size(NA)
```

Some functions have an inbuild option to remove NA values:
```{r}
x <- c(1, 2, NA, 4)
var(x)
var(x, na.rm = TRUE)
```


Let's get back to the lecture slides




\newpage

## (Long) Session 3: Vectors


### Type/ Mode

R automatically decides the mode (what type of thing the container contains), often easy:

```{r}
a <- c(1, 2, 3, 4)
a
class(a)
typeof(a)
```

```{r}
b <- c("a", "bunch", "of", "strings")
b
typeof(b)
```

```{r}
c <- c(TRUE, FALSE, T, T, F) # TRUE is same as T, FALSE is same as F
c
typeof(c)
```

and sometimes hard:

```{r}
d <- c(4, "word", T)
typeof(d)
d
```


we can force behaviours using as.<*> functions:

```{r, error = FALSE}
as.numeric(d)
as.logical(d)
```

Sometimes, conviently, R knows how to interpret text. 

```{r}
as.numeric(c("1", "two", "3", 4))
```

but R does not 'know' english, and does recognize numbers!Last, note the difference between warnings and errors! Here we get warnings, R does not break down.



### Object properties:

objects have properties, e.g.  a vector has a length:

```{r}
length(a)
```

you can access a particular element using `[]`

```{r}
a[1] 
```

each element in a vector can be given a name
```{r}
names(a) # NULL first!
names(a) <- b
a
```

Some useful ways of making vectors:
```{r}
e <- character(10) # creates an 'empty' vector of type character, with space reserved for 100 elements
e[5] <- "the fifth entry"
```


```{r}
f <- 1:10
g <- -5:3
h <- seq(0, 1, by = 0.1)
```

```{r}
class(f)
class(h)
```

actually there are multiple 'vector' classes: for each type of mode, a unique vector class exists (but all containing just one mode).


There is another property called 'dim': the dimensionality of the vector
```{r}
class(f)
dim(f)
dim(f) <- c(2, 5)
class(f)
```


### Vectorized Functions


Some functions can be applied to vectors:

```{r}
a * 2
a
a * a
g <- 1:3
g * 3
1:3 * 2
```

We called this 'vectorized' functions, they are usually superfast.


Some functions require vectors:

```{r}
mean(a)
```

Some functions behaviour depends on whether argument is vector or not:

```{r}
diag(3) # diagonal matrix
diag(c(1, 2, 3))
```

The behaviour of functions in R depends on the arguments you give it. R uses the information about the argument (the class) to determine which behaviour to perform! Be aware! Often only a single class may be used.


### IMPORTANT: Quirks and Perks: Vector Recycling!

```{r}
a <- 1:2
b <- 1:3
a * b
```

Note how the first element of `a` is used for a second time to multiply it with the 3rd element in `b`!




\newpage

## Session 4:


### Indexing:

Let us create a simple vector as an example

```{r}
x <- 1:10
x
```

Let us give each of the entries a name

```
names(x) <- letters[1:10]
x
```

```{r}
x[1]
x["b"]
x[-1]
x[c(T, F, F, F, F, F, F, F, F, F)]
```


What happens if we do:
```{r, error = FALSE}
x[0]
x[11]
x["m"]
```

Mind blowing! The Expansive power of vector recycling: 

```{r}
rep(c(1,2), each = 10)
Booleans <- c(T,F)
Booleans[rep(c(1,2), each = 10)]
```



### filtering

Using logical operators we can make R check if a particular condition is TRUE or FALSE

```{r}
1 < 2
2 < 1
```


Using logical operators applied to vectors we can automatically created vectors with corresponding TRUE and FALSE that check, elementwise, if the condition holds for the elements in x

```{r}
x > 5
x[x > 5]
```



\newpage 

## Session 5:


### Workspace

Usually your script is enough to reproduce an entire workspace    

Sometimes analyses will take a very long time and saving the script only can be very inconvenient

```{r}
ls()
save(list=ls(), file="my_workspace.Rdata") # this stores your workspace 'somewhere' on your PC, we'll look later at how to tell R what 'somewhere' to choose
rm(list=ls()) # remove everything
# now let's load everything back in
load("my_workspace.Rdata")
```


If you really want to save everything you've ever done:

```{r, eval = FALSE}
savehistory(file = ".Rhistory")
```



### Packages:

base is a package loaded into R by default

```{r}
library("base")
```

R won't give you any feedback unless something may have gone wrong. For anay help on a library, or function:


```{r, eval = FALSE}
library(help="base")
?sample
```

a typical library you might use that is not loaded by default, but installed by default

```{r}
library("foreign")
library("MASS")
```


a new library that you will need to use for extra exercises at `home'

```{r, eval = FALSE}
install.packages("swirl")
library("swirl")
rm(list=ls())
swirl()
```



