---
title: "SCR assignment 1"
output: pdf_document
---




```{r, include = FALSE}
load("0_data/scr_exam_1_the-basics.Rdata")
```

## Instructions / Preparatory code

Make sure that your write your `R` code in the `R` code chunks. When you wish or need to write text to answer one of the questions, don't use `R` comments, but just write the text outside the `R` code chunk. 

Small comments to clarify the `R` code should of course appear inside the `R` code chunks.



## 1.1 Reading in the data

### a

```{r}
```

### b

```{r}
```

### c

```{r}
```


## 1.2 Exploratory 

### a.

```{r}
```

### b.

```{r}
```

### c.

```{r}
```

### d.

```{r}
```

### e.

```{r}
```


### f.

```{r, researcher_age_mistake, eval=F}
```

## 1.3 Working with dates

### a.

```{r}
```

### b.

```{r}
```

### c.

```{r}
```

## 1.4 Doing the work

### a.

#### i.

```{r}
```

#### ii.

```{r}
```

### b.

### b: The short route

```{r}
```


### b: or The long route

#### i

```{r}
```

#### ii

```{r}
```

#### iii

```{r}
```

#### iv

```{r}
```

#### v

```{r}
```

### c

```{r}
```

### d

```{r}
```

### e

```{r}
```
