---
title: "SCR Week 3: live coding"
output: 
    pdf_document:
      keep_tex: yes
      fig_caption: yes
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{placeins}
---


# More about Functions

## Scoping | Session 1

```{r, eval=FALSE}
numerator_value <- 12

GiveRemainder <- function(divisor) {
   return(numerator_value %% divisor)
}
GiveRemainder(10)
numerator_value

# numerator_value <- 12
GiveRemainder(10)
```


## Setting a value | Session 2

This is part of an Object Oriented Programming style

```{r}
person <- list(name = "Who?", age = 34, gender = "male")
person[["name"]]
```

```{r}
GetName <- function(x) {
  return(x[['name']])
}
GetName(person)
```

```{r, error = TRUE}
GetName(person) <- ""
```

However, apparantly in R, we can assign values based on a function call.These are functions as `names()`, `class()`, `mode()`... 

```{r}
obj <- c(a = 2, b  = 3, c = 3)
names(obj)
names(obj) <- paste("V", 1:3, sep = "")
names(obj)
```


The trick on how to do this... (will be skipped if not enough time!)

```{r}
MyName <- 'MyName<-' <- function(x, value = x[['name']]){
  x[['name']] <- value
  return(x)
}
```

```{r}
MyName(person) <- ""
MyName(person)
```



## Parts of a function | Session 3

```{r}
Multiply <- function(){}
```

```{r}
args(Multiply)
formals(Multiply)
```


```{r}
Multiply <- function(x, y = x, ...){
  
  args <- list(...) # remember swirl?
  
  if (!is.null(args$message)){ # remember tries to find a named entry / argument?
    print(args$message)
  }
  
  return_value <- x * y
  return(return_value)
}
```


```{r}
args(Multiply)
formals(Multiply)
```


```{r}
Multiply(2)
Multiply(2, 4)
Multiply(x = 2, y = 4, message = "This is a multiplication of 2 and 4")
```




### Down the rabbit hole

When getting help on a function, don't forget to look at the examples at the end of the file.

```{r, eval=FALSE}
?mean
example(mean) # or just call them.
```

Other options: 

```{r, eval = FALSE}
View(Multiply)
Multiply <- edit(Multiply)
Multiply(2,3, message ="hi")
```


R is open source! 

```{r, eval = FALSE}
View(apply) # oh dear
```

Something simpler then?

```{r, eval = FALSE}
View(mean)
methods("mean")
View(mean.default)
?.Internal
.Internal(mean(1:100))
View(.Internal)
.Primitive(".Internal")
```

To really find out how the function `mean()` works, download the source of R (e.g. not the binary versions...). Then, find the file: names.c, and look for "mean", and find "do_summary". So, we have to go to the `summary.c` file and look for do_summary...?  pff

Can we find out the source code for `rnorm()`...? 

```{r, eval = FALSE}
View(rnorm) # this is new ?
``` 

`.Call?` Darn it! We have to go -> names.c -> etc.. 

Whenever you encounter `C()`, `.Call()`, `.Fortran()`, `.External()`, or `.Internal()` and `.Primitive()`,  don't bother (for this course!)




\newpage 

##  Factors | Session 4

```{r}
set.seed(20180928)
my_vector <- letters[sample(1:10, 25, replace = T)]
class(my_vector) # works like a names function 
```

Creating a factor

```{r}
my_factor <- factor(my_vector)
my_factor <- factor(my_vector, levels = letters[1:10])
attributes(my_factor)
levels(my_factor) # works like a names function
```

Nice feature: factors protect you against invalid entries

```{r, error = TRUE}
my_factor[1] <- "z"
```

Removing factor

```{r}
as.character(my_factor)
unclass(my_factor)
attributes(my_factor) <- NULL  # works like a names function
my_factor
```


More ways to create factors

```{r}
my_factor <- factor(my_vector, levels = unique(my_vector))
my_factor
```

Also check out `gl()`...

The ordering in the levels, determines the order in the factor:

```{r}
my_factor <- factor(my_vector, levels = letters) 
my_factor[1] <- "z"
```

```{r}
my_factor <- factor(my_vector, levels = rev(letters)) 
my_factor[1]
# unclass(my_factor)[1]
```

```{r}
my_factor <- factor(my_vector, levels = letters, labels = rev(LETTERS))
my_factor <- factor(my_vector, levels = letters)
my_factor
```


Removing unobserved levels:

```{r}
my_factor <- droplevels(my_factor)
```


Quirks and perks? factor is 'a kind of' container... 

```{r}
class(my_factor)
mode(my_factor) # ???
my_factor + 1 # pfew...
as.numeric(my_factor) # ok...:s
as.numeric(factor(my_factor, levels = rev(letters))) # you know why?
```



## table | Session 5

```{r, eval = FALSE}
N <- 3e4

my_numbers <- sample(rep(1:3, each= N / 3))
table(my_numbers)
my_chars <- sample(letters[1:3], 3e4, replace=T)
table(my_chars)

table(my_numbers, my_chars)

my_data <- data.frame(my_numbers, my_chars)
table(my_data)

my_logicals <- sample(c(TRUE, FALSE), N, replace=T)
table(my_numbers, my_chars, my_logicals)

my_data <- cbind(my_data, my_logicals)
table(my_data)
```


## arrays | Session 6

```{r, eval = FALSE}
my_array <- array(0, dim = c(2, 3))
my_array
class(my_array)
```

```{r}
my_array <- array(1:24, dim = c(4, 3, 2))
```

Remember how data is filled up over the dimensions.

```{r}
class(my_array)
my_array
```

```{r}
my_array[1, 1 , 1]
my_array[, , 1]
my_array[1, , ]
my_array[, 1, ]
```

```{r}
my_array <- array(1:120, dim = c(5, 4, 3, 2))
# pff...
```

The table one was much easier right? Everything is mode numeric and we can  turn it into proportions easily:

```{r, eval = FALSE}
my_table <- table(my_data)
my_table/ sum(my_table)
```


### `apply()`:

```{r}
apply(my_table, 3, sum) # row marginals
apply(my_table, 1, sum) # depth(?) marginals
apply(my_table, c(1,2), sum) # row / column marginals (i.e. summing the two layers)
```



### `tapply()`?

```{r}
med_example <- data.frame(
  patient = 1:100,
  age = rnorm(100, mean = 60, sd = 12),
  treatment = factor(
    rep(c(1,2), 50), 
    labels = c("treatm", "control")
  ),
  hospital = factor(
    rep(c(1,2), each = 50), 
    labels = c("A", "B")
  )
)
```

```{r}
set.seed(42)
med_example$QL <- 30 + 10*(med_example$hospital == "A") + 30*(med_example$treatment == "treatm") + rnorm(100, mean = 10, sd = 10)
```

```{r}
output_tapply <- tapply(
  X = med_example$QL, 
  INDEX = list(
    med_example$hospital,
    med_example$treatment
  ), 
  FUN = mean
)
output_tapply
```



```{r, eval = FALSE}
aggregate(
  QL ~ hospital + treatment, 
  FUN = mean,
  data = med_example
)
```





