---
title: "SCR Exam 1: the basics"
output: pdf_document
---

```{r, include = FALSE}
load("0_data/scr_exam_1_the-basics.Rdata")
```

## Instructions / Preparatory code

Make sure that your write your `R` code in the `R` code chunks. When you wish or need to write text to answer one of the questions, don't use `R` comments, but just write the text outside the `R` code chunk. 

Small comments to clarify the `R` code should of course appear inside the `R` code chunks.



## 1 Introduction

### a

```{r}
```

### b

```{r}
```

### c

```{r}
```

## 2 Set-up

### a

```{r}
```

### b

```{r}
```


## 3 Random imputation

### a

```{r}
```

### b

```{r}
```

### c

```{r}
```

### d

```{r}
```

### e

```{r}
```

### f

```{r}
```

## 4 Informative imputation

### a

```{r}
```

### b

```{r}
```

### c

```{r}
```

### d

```{r}
```

### e

```{r}
```

## 5 Impute again, analyze again

### a

```{r}
```

### b

```{r}
```

### c

```{r}
```


## 6 Self Service Analytics

### a

```{r}
```


### Bonus

```{r}
```
