---
title: "Monte-Carlo Studies"
subtitle: "SCR Lecture 7"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "R-team"
output:
  ioslides_presentation:
    keep_md: yes
    fig_caption: true
    fig_height: 4
    fig_width: 5
    transition: faster
    css: 0_css/SCR1718.css
    logo: 0_css/logo.png
    template: 0_template/R1718style.html
    highlight: tango
    widescreen: TRUE
    mode: standalone
editor_options: 
  chunk_output_type: console
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
- \usepackage{placeins}
- \usepackage{amsmath}
---


```{r label = setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  dev.args = list(bg = 'transparent'),
  fig.align = "center",
  cache = FALSE
)
```

# SCR Part 2

## SCR Part 2: 
 
- Statistical Computing: (Re)sampling studies (4 lectures)

- Reproducibility: Dynamic Documents / Tidyverse (2 Lectures)            


## About the assignment and the exam

- ~~Midterm~~

- Assignment     

- Next Exam    


# Monte-Carlo Studies

##  Outline of today

Using Monte-Carlo experiments to study a certain statistic. 

Chapter 6 Book Rizzo!

Objectives: 

* Get familiar with Statistical Jargon in Monte Carlo studies
* Awareness of simulation error
* A general framework for Monte Carlo studies to e.g.:
    - approximate the bias of an estimator
    - approximate the type I error and/or Power of a test



## Monte Carlo Studies

We know $f_X$, and we need to study $T(\mathbf{X})$.

DISTRIBUTION: $f_X$

DATA: $\mathbf{X} = \{ X_1, \ldots, X_n \} \overset{\text{iid}}{\sim} f_X$

STATISTIC: $T(\mathbf{X})$     


## Why? 

In statistics / data science:

* to investigate statistical properties (e.g. the properties of a sampling distribution);
* to evaluate the performance of statistical procedure (e.g. thesis!);
* to clarify or verify your math (e.g. see the central limit theorem at work)


## All about built-in Random Variate Generators 

Functions in `R` to generate variates randomly:

* `rnorm()`: Normal distribution
* `rbinom()`: Binomial distribution
* `rt()`: t-distribution
* `runif()`: Uniform distribution
* `rchisq()`: Chi-squared distribution
* `rgamma()`: Gamma distribution
* `rexp()`: Exponential distribution


##  Monte Carlo Studies: Today

We know $f_X$, and we need to study $T(\mathbf{X})$.

For example to estimate distributions, integrals, probabilities: 

* Sampling distribution    
* Bias of an estimator   
* Mean Squared Error (MSE)   
* Simulation Error   
* Type I error   
* Power   

<!-- Goal is to get started with an example, then formalize some of it -->


# A first example: throwing a ball


## Throwing a ball

Suppose you ask me to throw a ball 50 meters, how far will my throw on average be?

How would you answer this question?

<!-- drawing on the board -->


# Exercises part 1


## Throwing a ball: summary

Which answer did you prefer, the analytical one or the monte carlo one?

<!-- Which one is faster? In more difficult cases? If you are trained to program these? -->

The answer you got differed from the answer of your neighbour, why?

<!-- briefly about simulation error -->

The exercise was about different sample and sampling distributions, which ones?

<!--  - Sample distribution of throws (throwing 10 times), 
      - Sampling distribution of average of throws (1000 times, average of 10 throws)
      - Sampling distribution of average of average of throws (last comparison between you and fellow student) -->


# Sampling Distribution 


## Sampling distribution

<!-- Will for example help us think about the properties of a test -->

DATA: $\mathbf{X} = \{ X_1, \ldots, X_n \} \overset{\text{iid}}{\sim} f_X$

STATISTIC: $T(\mathbf{X})$     

The distribution of $T$ is called sampling distribution 


## Theoretical Sampling Distribution

Towards a theoretical sampling distribution:

If $\mathbf{X} = X_1, \ldots, X_n \stackrel{i.i.d.}{\sim} N(0, \sigma^2)$    

with sample variance $S^2 = \frac{1}{n - 1} \sum^{n}_{i = 1} (X_i - \overline{X})^2$,    

then 

$$T(\mathbf{X}) = \frac{\overline{X}\sqrt{n}}{S} ~ \sim t_{n - 1},$$

where $t_{n - 1}$ is a theoretical $t$-distribution.
<!-- Proof: http://stats.stackexchange.com/questions/151854/a-normal-divided-by-the-sqrt-chi2s-s-gives-you-a-t-distribution-proof  -->

## Estimation of a Sampling distribution

For $b = 1,\ldots,B$

* generate replicate $b$ for data $\mathbf{X}$, called $\mathbf{X}^{b}$
* compute $T^b =T(\mathbf{X}^{b})$   

Make a plot (histogram, density, ecdf) of $T^1,\ldots,T^B$.


## Live coding: monte carlo sampling distribution

```{r}
set.seed(20171102)
N <- 25
B <- 10000

T_function <- function(X){
  n <- length(X)
  return(mean(X) * sqrt(n) / sd(X))
}

T_replicates <- numeric(B)
for (b in 1:B){
  X <- rnorm(N)
  T_replicates[b] <- T_function(X)
}
```

## Live coding: monte carlo sampling distribution
```{r}
hist(T_replicates, freq=F, breaks="FD", col='lightgreen')
x_seq <- seq(-10, 10, by=0.01)
lines(x_seq, dt(x_seq, df=N-1), lwd=2)
```

# Standard Error 

## Standard Error

DATA: $\mathbf{X} = \{ X_1, \ldots, X_n \} \overset{\text{iid}}{\sim} f_X$

STATISTIC: $T(\mathbf{X})$     

The standard deviation in the sampling distribution of $T$ is called the standard error.   

## Example: Standard Error 

<!-- Will for example help us think about the properties of a test -->

If $\mathbf{X} = X_1, \ldots, X_n \overset{\text{iid}}{\sim} N(0, \sigma^2)$    

and $T(\mathbf{X}) = \overline{X}$,    

then standard error is $$\sigma_{\overline{X}} = \frac{\sigma}{\sqrt{n}}$$      

(Also known as S.E., SE)        


## Example: monte carlo Standard Error

DATA $\mathbf{X} = X_1, \ldots, X_n \overset{\text{iid}}{\sim} N(0, \sigma^2)$    

STATISTIC $T = T(\mathbf{X})$   

DETERMINATION OF STANDARD ERROR BY SIMULATION:

For $b = 1,\ldots,B$

* generate independent replicate $\mathbf{X}^b$    
* compute $T^b =T(\mathbf{X}^b)$   

Compute root of variance $T^1, \ldots, T^B$, i.e.

$$
\widehat{\text{SE}_T} = \sqrt{ \frac{1}{B} \sum^B_{b = 1} (T^b - \overline{T})^2}, \qquad \qquad \textrm{where } \overline{T} = \frac{1}{B} \sum^B_{b = 1} T^b
$$


## Live coding: standard error, monte carlo in `R`

```{r}
N <- 100
B <- 1000

sample <- rnorm(N)
mean(sample)

#theoretical:
1/sqrt(N)
#'empirical':
sd(replicate(B, mean(rnorm(N))))
```

# Exercises part 2


# Bias and MSE of estimators

## Bias of estimators

ESTIMATOR $T = T(\mathbf{X})$ OF QUANTITY $\theta$   

BIAS of $T$: $\textrm{E}\left[T\right] - \theta$

EXAMPLE OF THEORETICAL BIAS:   

- If $X_1, \ldots, X_n \stackrel{i.i.d.}{\sim} N(0,\sigma^2)$, then bias of $\bar{X}$ for $\theta$ is 0.   


## Bias of estimators

ESTIMATOR $T = T(\mathbf{X})$ OF QUANTITY $\theta$   

BIAS of $T$: $\textrm{E}\left[T\right] - \theta$

MONTE CARLO EXAMPLE OF THEORETICAL BIAS:   

For $b = 1,\ldots,B$

* generate independent replicate $\mathb{X}^b$ 
* compute $T^b =T(\mathbf{X}^b)$   

Compute:    
$$
\widehat{(\textrm{E}\left[T\right] - \theta)} = \frac{1}{B} \sum^B_{b = 1} T^b - \theta
$$


## MSE: Mean Square Error

MSE of an estimator is the sum of its Bias$^2$ and its Variance

<!-- Dartboard drawing -->

Recall throwing a ball: the mean throw was already off.

So the average squared error (from target to throwing distance) will be off by bias^2, plus the variance around the throw


## MSE: Mean Square Error

ESTIMATOR $T = T(\mathbf{X})$ OF QUANTITY $\theta$    

The *mean square error* (MSE) of $T$ is $\textrm{E}\left[(T - \theta)^2\right]$.

## MSE: decomposed

MSE:     
$\textrm{E}[(T - \theta)^2] = \textrm{E}[T^2] + \theta^2 - 2\textrm{E}[T]\theta$     

Squared bias:    
$(\textrm{E}[(T - \theta)])^2 = (\textrm{E}[T])^2 + \theta^2 - 2\textrm{E}[T]\theta$    

Variance:    
$\textrm{E}[(T - \textrm{E}[T])^2 = \textrm{E}[T^2] - ( \textrm{E}[T])^2$   
   
MSE = Squared bias + Variance

## Verify Theoretical MSE  

If $\mathbf{X} = X_1, \ldots, X_n \overset{\text{iid}}{\sim} N(\theta, \sigma^2)$, then MSE of $\overline{X}$ for $\theta$ is \(\sigma^2/n\).

USING MONTE CARLO APPROXIMATION:

For $b = 1,\ldots,B$   

* generate independent replicate $\mathbf{X}^b$
* compute $T^b =T(X^b)$   

Compute 
$$
\widehat{MSE} = \frac{1}{B} \sum^B_{b = 1} (T^b - \theta)^2
$$
<!-- The estimator may be unbiased, but the estimate itself will most definately be biased! -->


# Exercises part 3

## Levels of sampling.

Be very mindful of the level at which you need extra repetitions:

* If you want a better estimate of the median per sample, increase $n$
* If you want a better estimate of the mean of the median, increase $B$ (to decrease simulation error)
* If you want a better estimate of the standard error of the sampling distribution of the mean of the median, repeat the expirment with $B$ replicates $M$ time, and check out the average of the $M$ standard errors...

## Simulation error

By making $B$ larger, the simulation error can be made arbitrarily small. $B = 10,000$ is desirable, but $B = 1,000$ is typical, and only $B = 100$ may be feasible.

If $\textrm{E}\left[T\right]$ is estimated by $\overline{T}$ for i.i.d. $T^1,\ldots,T^B,$ then CLT kicks in (again):

$$
\bar{T}_B - \textrm{E}\left[T\right] \sim N(0, \sigma_T^2/B).
$$

Why would CLT kick in?

Thus the standard error of simulation, $\sqrt{\textrm{var}_T/ B},$ and can be estimated by

$$
\sqrt{B^{-1} \sum^B_{b=1} (T^b - \overline{T})^2/B }
$$

## Standard error of Simulation

By making $B$ larger, the simulation error can be made arbitrarily small. $B = 10,000$ is desirable, but $B = 1,000$ is typical, and only $B = 100$ may be feasible.

EXAMPLE:   

If the MSE, $\textrm{E}\left[(T - \theta)^2\right]$, is estimated by $\widehat{MSE} = B^{-1} \sum^B_{b = 1}(T^b - \theta)^2$, then the simulation error can be estimated by 
$$
\widehat{\text{SE}_\text{sim}} = \frac{1}{B}\sqrt{\sum^B_{b=1}\left[ (T^b - \theta)^2 - \widehat{MSE} \right]^2}
$$

<!-- [ The leading fraction 1/B (and not 1/√B) is not a mistake. Do you see where it came from?] Compare with previous slide! -->

# Size / Type 1 error

## Size / Type 1 error

STATISTIC $T = T(\mathbf{X})$    

$H_0 :=$ null hyptohesis

Reject $H_0$ if $T \in K_0$.     

$K_0: =$ domain of values for $T$ which are unlikely under $H_0$.

$\alpha = P_{H_0}(T \in K_0)$.   

Remember: Most tests are constructed so that the size equals the level (= \alpha), e.g. 0.05.


## Compute size or Type 1 error by Simulation.

Reject $H_0$ if $T \in K_0$.     

$K: =$ domain of values for $T$ which are unlikely under $H_0$.

What is $\alpha = P_{H_0}(T \in K)$? 

For $b = 1,\ldots,B$   

* generate independent replicate $\mathbf{X}^b$ using the null distribution  
* compute $T^b =T(X^b)$   
* compute

$$
\hat{\alpha} = \frac{1}{B} \sum^B_{b = 1} 1_{T^b \in K} = \textrm{fraction rejections},
$$

where $1_{T^b \in K} = 1$ if $T^b \in K$, and $0$ otherwise.

<!--  
If H0 is composite, must simulate using the worst case null distribution, or repeatedly simulate using all null distributions and take the maximum of the simulated $\alpha$.
-->


## Live coding: Size / Type 1 error in `R`

Suppose we define the following. We know that the average of a sample is distributed according to a normal distribution, with standard deviation $1$. The *null hypothesis* is that the average of this distribution is $0$. The test, for $N=10$ is: we reject the null hypothesis that the mean of the population is $0$ if the test statistic is smaller than -1, or larger than 1.

```{r}
set.seed(21135)
N <- 10
B <- 1000

# simulate from null distribution.
z <- replicate(B, rnorm(N))

t_in_k <- (z < -1) + (z > 1)
mean(t_in_k)
```

With this test, we falsely reject the null hypothesis for 31.54% in samples of size $N=10$.

# Power

## Power

STATISTIC $T = T(\mathbf{X})$    

$H_1 :=$ Alternative hypothesis

STATISTIC $T = T(\mathbf{X})$ 

We reject $H_0$ if it falls in critical region $K_0$

The power is $P_{H_1}(T \in K_0)$ viewed as function of the parameter space in $H_1$.   


## Power

The power is $P_{\theta_1}(T \in K_0)$ viewed as function of the alternative $\theta_1$ from $H_1$.   
 
DETERMINATION OF POWER BY SIMULATION

For $b = 1,\ldots,B$

* generate independent replicate $\mathbf{X}^b$ while *using alternative* $\theta_1$    
* compute $T^b =T(\mathbf{X}^b)$   

Compute 

$$
\frac{1}{B}\sum^B_{b = 1} 1_{T^b \in K_0}
$$

Repeat for all alternatives.

## When you do **Simulations / Monte Carlo Experiments**

Note to self: 

* Good documentation is essential! 
* Write separate programs for each case or keep a precise record 
* Think about which output you need to save, and do that in a structured way    
* Make programs 're-startable' in case of emergency..
* Divide the work over more computers.     

## Recap of today

* Statistical distribution functions 
* Sampling distribution
* Estimators: Bias and MSE
* Standard Error of Simulation
* Type 1 error
* Power

# Work on the exercises / self-study | and see you next week!

